\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-24 NAR_DRAFT.tex
citations and things

\batcave is an algorithm that leverages the biology of individual tumor mutation profiles to improve identification of low allelic frequency somatic variants.
Our current implementation is built on MuTect, one of the most widely used somatic variant callers.
\batcave improves on the classification accuracy of MuTect in synthetic data (Fig.~\ref{NAR-wes_fig}A-D and Table~\ref{table:01}) across the entire range of recall and specificity.
Moreover, \batcave is better calibrated than MuTect at relevant posterior probability thresholds (\ref{NAR-ppv_fig}), allowing researchers and clinicians to make informed choices about the trade-off between precision and recall.
For real data, testing on validated calls shows that \batcave does not degrade performance for variants that are relatively easy to identify (Fig.~\ref{NAR-wes_fig}E\&F).
The \batcave algorithm can thus be included in a wide variety of sequencing pipelines.

The \batcave algorithm incorporates genomic context into the probabilistic model for variant calling.
Our current implementation focuses on trinucleotide context, which is known to have a large effect on local mutation rates \cite{Martincorena2015,Hollstein2017}.
There are, however, many other aspects of genomic context that can affect local mutation rates \citep{Buisson2019}, including replication timing \citep{Stamatoyannopoulos2009}, expression level \citep{Pleasance2010}, and chromatin organization \citep{Schuster-Bockler2012}. 
\rngcomment{Move individual citations to the factors they support.}
These additional forms of context could also be incorporated into the \batcave mutation prior.
In fact, they may be simpler to incorporate than trinucleotide context, because their effect on the prior may vary less among tumors. \rngcomment{Can we back this assertion up?} 
In the long run, we believe that incorporating more tumor biology into variant calling models will often lead to improved performance.

\batcave divides the data into two classes: high- and low-confidence variants.
The high-confidence variants are used to estimate the mutational prior, which is then used to improve the calling of low-confidence variants.
Statistically, this is an empirical Bayesian approach~\cite{XXX}. \rngcomment{Add citations, and sharpen discussion if necessary.}
In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.
A key implicit assumption of our approach is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\cite{TrackSigpaper}, although those differences are relatively small.
A potential extension of the \batcave algorithm is process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
This approach might increasing sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.

>>>>

\batcave is an algorithm that leverages the biology of individual tumor mutation profiles to improve identification of low allelic frequency somatic variants.
Our implementation is built on MuTect, one of the most widely used somatic variant callers.
\batcave improves on the classification accuracy of MuTect in synthetic data (Fig.~\ref{NAR-wes_fig}A-D and Table~\ref{table:01}) across the entire range of recall and specificity.
Moreover, \batcave is better calibrated than MuTect at relevant posterior probability thresholds (\ref{NAR-ppv_fig}), allowing researchers and clinicians to make informed choices about the trade-off between precision and recall.
For real data, testing on validated calls shows that \batcave does not degrade performance for variants that are relatively easy to identify (Fig.~\ref{NAR-wes_fig}E\&F).
The \batcave algorithm can thus be included in a wide variety of sequencing pipelines.

The \batcave algorithm incorporates genomic context into the probabilistic model for variant calling.
Our current implementation focuses on trinucleotide context, which is known to have a large effect on local mutation rates \cite{Martincorena2015,Hollstein2017}.
There are, however, many other aspects of genomic context that can affect local mutation rates \citep{Buisson2019}, including replication timing \citep{Stamatoyannopoulos2009}, expression level \citep{Pleasance2010}, and chromatin organization \citep{Schuster-Bockler2012}. 
Some of these, such as replication timing and chromatin organization, could be incorporated into the \batcave mutation prior by observing the empirical distribution of mutations in the human germline \cite{Hodgkinson2011}.
Others, such as expression level, could be tumor-specific, but would require information not available in the variant calls to compute.
In the long run, we believe that incorporating more tumor biology into variant calling models will often lead to improved performance.

\batcave divides the data into two classes: high- and low-confidence variants.
The high-confidence variants are used to estimate the mutational prior, which is then used to improve the calling of low-confidence variants.
Statistically, this is an empirical Bayesian approach~\citep{Robbins1954}. \rngcomment{Add citations, and sharpen discussion if necessary.}
In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.
A key implicit assumption of our approach is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\citep{Rubanova2018a}, although those differences are relatively small.
A potential extension of the \batcave algorithm is process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
This approach might increasing sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.
